# Reproducible Research - Peer Assessment 1

## Load needed libraries
```{r}
library(ggplot2)
```

## Load and preprocess the data
0) The data file should be manually extracted from the zip file. 

1) Then, load / read the data.
```{r readData, echo=TRUE}
df<-read.csv("activity.csv", 
        header = TRUE,
        colClasses = c("numeric", "Date", "integer"),
        nrows = 17700)
```

2) Check that data is in useable format for analysis and tidy.
```{r checkFile, echo=TRUE}
str(df)
summary(df)
head(df)
tail(df)
```

## What is mean total number of steps taken per day?
Sum steps by days. Ignore missing/NA values. This is assumed to be the mean of the total number of steps for each day.
```{r SumStepsByDay}
stepsbyDate <- aggregate(x = df$steps, 
          by = list("date" = df$date), 
          na.rm = TRUE,
          FUN = sum, 
          simplify = TRUE)

names(stepsbyDate)[2] <- "steps"   #rename x from aggregate() to steps

head(stepsbyDate)
```

1) Create a histogram of total number of steps taken each day.  
```{r TotalStepsByDay}
qplot(steps, data=stepsbyDate, binwidth = 1000)
```

2) Calculate the mean and median.
```{r}
meanSteps <- mean(na.omit(stepsbyDate$steps))
medSteps <- median(na.omit(stepsbyDate$steps))
```
Mean of steps taken per day:  `r format(meanSteps, nsmall = 1)`  
Median of steps taken per day: `r format(medSteps, nsmall = 1)`

## What is the average daily activity pattern?
Average the steps for all days in the original data set by the 5-minute intervals

```{r}
stepsbyInterval <- aggregate(x = df$steps, 
          by = list("interval" = df$interval), 
          na.rm = TRUE,
          FUN = mean, 
          simplify = TRUE)

names(stepsbyInterval)[2] <- "steps"   #rename x from aggregate() to steps
str(stepsbyInterval)
head(stepsbyInterval)
```

1) Create a time series plot of average-number-of-steps-for-all-days per interval.  Because the x-asis is continous and the time interval data ranges from XX:00 to XX:55, there is a 45 minute gap in each hour from 00:00 to 23:55. This approach is consistent with the project example.
```{r AverageDailyPattern}
qplot(interval, steps, data=stepsbyInterval, geom="line")
```

2) Identify the 5-minute interval with the maximum average-number-of-steps-for-all-days
```{r}
maxInterval <- stepsbyInterval[which.max(stepsbyInterval[,2]),]
```

The interval with the maximum, on average, number of steps is `r format(maxInterval[1], nsmall = 0)`.

## Imputing missing values
1) Identify the total number of missing values.
```{r}
nacount <- sum(is.na(df$steps))
```
The total number of rows with NAs is `r format(nacount, nsmall = 0)`

2) Create a method to fill in the missing values of steps. Will use the average-for-all-days of the 5-minute interval.
```{r}
# Use a logical vector to identify which rows in $steps contain NA
nasteps <- is.na(df$steps)
# create a vector of length(nasteps) that contains the  interval average by repeating the interval average vector by the number of days in the dataset
intavgsteps <- rep(stepsbyInterval$steps, length(stepsbyDate$steps))
```

3) Create new dataset equal to original but with NAs replaced by average-for-all-days of the 5-minute interval.
```{r}
#replace NA's in df$steps with interval averages
newdf <- df
newdf$steps[nasteps] <- intavgsteps[nasteps]
head(newdf)        
```

4a) Create a histogram of the total steps taken each day using the new dataset (i.e., NAs replaced by averages). 
```{r TotalStepsReplacedNAs}
newstepsbyDate <- aggregate(x = newdf$steps, 
          by = list("date" = newdf$date), 
          na.rm = TRUE,
          FUN = sum, 
          simplify = TRUE)

names(newstepsbyDate)[2] <- "steps"   #rename x from aggregate() to steps
str(newstepsbyDate)
```

```{r}
qplot(steps, data=newstepsbyDate, binwidth = 1000)
```

4b) Report the mean and median of steps taken per day.
```{r}
newmeanSteps <- mean(na.omit(newstepsbyDate$steps))
newmedSteps <- median(na.omit(newstepsbyDate$steps))
```
Mean of steps taken per day (new):  `r format(newmeanSteps, nsmall = 1)`  
Median of steps taken per day (new): `r format(newmedSteps, nsmall = 1)`

4c) Do the mean and median of steps, with the NAs replaced, differ from the original data set?

Subtracting the new mean and median from the original calculations gives the following differences:  
Mean difference: `r format(newmeanSteps - meanSteps, nsmall = 1)`  
Median difference: `r format(newmedSteps - medSteps, nsmall = 1)`

4d) The impact of replacing NAs with estimates is that the mean and median increase.


## Are there differences in activity patterns between weekdays and weekends?

1) Using the new data set (no NAs), create a factor variable with two values - a) weekday, b) weekend.

```{r}
newdf$weekday <- weekdays(newdf$date, abbreviate = TRUE)

newdf$dayend <-as.factor(ifelse (newdf$weekday %in% c("Sat","Sun"), "weekend","weekday"))
head(newdf)
tail(newdf)
newdf[2000:2005,]
```

2) Make a panel time-series plot averaged across all weekday and weekend days. 

Aggregate the new data-set by interval and dayend.
```{r}
newstepsbyInterval <- aggregate(x = newdf$steps, 
          by = list("dayend" = newdf$dayend,
                    "interval" = newdf$interval),
          na.rm = TRUE,
          FUN = mean, 
          simplify = TRUE)

names(newstepsbyInterval)[3] <- "steps"   #rename x from aggregate() to steps
```

Create a time series plot of average-number-of-steps-for-all-days per interval. Use ggplot2/ggplot.

```{r AverageStepsbyWeekendWeekday}
g <- ggplot(newstepsbyInterval, aes(interval, steps))
# ggplot2 facet strip on right
g + geom_line(color="blue") + 
        facet_grid(dayend ~ .) + 
        xlab("Interval") + 
        ylab("Number of Steps") + 
        theme_bw() + 
        theme(strip.background = element_rect(fill = 'beige'), 
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank() )
```

